Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-29 paper.tex
writing

Let's compute some singular values and vectors.  (And the the wrong answer—try again.)  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
where
$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
Presumably, the right singular vectors have the $f_n$ of least norm such that 
$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$

What is the appropriate norm to use?  In the case that $χ$ is a characteristic function on the domain $D$, it is presumably the truncated Bargmann norm
$$〈g,f〉=\int_D e^{-|α|²}g*(α)f(α)\,d²α=\int e^{-|α|²}χ(α)g*(α)f(α)\,d²α.$$
Note that this is only an inner product when $g$ and $f$ are restricted to $D$.  In this case, we wish to minimise 
$$‖f_n‖²=2π\int_0^b re^{-r²}f²_n(r)\,dr,$$
given 
$$2\sqrt π(n!)^{-½}\int_0^b r^{n+1}e^{-r²}f_n(r)dr=1.$$
If one considers changing $f$ by $Δa$ in an interval around $a$, the Lagrange condition is
$$re^{-r²}f_n(r)\propto r^{n+1}e^{-r²},$$
or 
$$f_n(r)\propto rⁿ.$$
The analytic functions still win.  Let $f_n(r)=rⁿ$.  Then
$$‖f_n‖²=2π\int_0^b r^{2n+1}e^{-r²}\,dr$$
and
$$‖ \Gx f_n‖=2\sqrt π(n!)^{-½}\int_0^b r^{2n+1}e^{-r²}dr,$$
so we have
$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{2\over n!}
	\left(\int_0^b r^{2n+1}e^{-r²}dr\right)^½.$$
According to Wolfram,
$$\int_0^b r^{2n+1}e^{-r²}dr=½\left(Γ(n+1)-Γ(n+1,b²)\right),$$
where $Γ$ is Mathematica's version of the incomplete gamma function.  Therefore, recalling that $n!=Γ(n+1)$,
$$σ_n=\left(1-{Γ(n+1,b²)\over Γ(n+1)}\right)^½.$$
Unfortunately, Matlab defines the incomplete gamma function in a different way than Mathematica does.  Fortunately, that Matlab function, {\tt gammainc}, is exactly the square of the right-hand side of the above equation.

This all seems wrong, because the incomplete gamma function varies the wrong way with $n$.  It becomes small as $n→0$.

Numerical experiments in the script {\tt sve.m} support this.  The operator $G$ was discretised on a grid in phase space, with spacing $h=0.3$, and various extents shown in the graphs.  Ket space was discretised by taking inner products with the number states $|0〉$ through $|10〉$.  The graph below shows the singular values of the discretised operator converging to $π^{-½}$, as the extent of the grid passes through $|α|²=10$.

@@

The following three figures show the right singular vectors, sampled versions of phase space functions.  The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.  These look very much like number states; that is an artefact of the way ket space was truncated and and phase space was discretised.  In the continuous limit there is only one singular value, and the vectors are degenerate.  For the $4×4$ grid, number states 5-10 don't fit, and the error in their singular values is of order 0.1 or more.  As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.

\topinsert \hbox{\XeTeXpicfile rsv2.pdf width 0.5\hsize
\XeTeXpicfile rsv3.pdf width 0.5\hsize}
\XeTeXpicfile rsv4.pdf width 0.5\hsize\hfil\break
The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase spaces is truncated to a square about the origin, whose size varies.  These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert

>>>>

Let's compute some singular values and vectors.  Suppose that $χ$ is the characteristic function of a disk, with
$$χ(α*)=\cases{1 & $|α|≤b$ \cr 0 & otherwise.}$$
From symmetry, let's guess that the singular kets are the number states—we will check this numerically later.  It can be shown by a Lagrange multiplier argument that the right singular functions are the truncations to the disk of $f_n=〈Γ⁺|n〉$.  Then the singular values are
$$σ_n={‖\Gx f_n‖\over ‖f_n‖}.$$
We take
$$f_n=π^{-½}〈α|n〉e^{½|α|²}=π^{-½}∑_{n=0}^∞ {α^{n\ast}\over \sqrt{n!}}.$$

First evaluate $‖f_n‖$, using the Bargmann norm.  We have
$$\eqalign{ ‖f_n‖²&={1\over π}\int_{|α|≤b} e^{-|α|²}{|α|^{2n}\over n!}\,d²α \cr
	&={1\over π}\int_0^b e^{-r²}{r^{2n}\over n!}2πr\,dr \cr
	&={1\over n!}\int_0^{b²} u^ne^{-u}\,du=Γ(n+1,b²),}$$
where $Γ(n,x)$ is the incomplete gamma function, as defined in Matlab.  (Everyone defines this a different way.)  The other norm is given by
$$\eqalign{ \Gx f_n&=π^{-½}\int_{|α|≤b} |α〉e^{-½|α|²}f_n(α*)\,d²α \cr
	&={1\over π}∑_{m=0}^∞\int e^{-|α|²}{α^{m}\over \sqrt{m!}}|m〉{α^{n\ast}\over \sqrt{n!}}\,d²α \cr
	&={1\over π}\int_0^b dr re^{-r²}∑_{m=0}^∞{r^{m+n}\over \sqrt{m!n!}}|m〉\int_0^{2π} dφe^{i(m-n)φ} \cr
	&=\int_0^b dr 2re^{-r²}{r^{2n}\over n!}|n〉 \cr
	&=Γ(n+1,b²)|n〉.}$$

The singular value comes out to 
$$σ_n={‖ \Gx f_n‖\over ‖f_n‖}=\sqrt{Γ(n+1,b²)}.$$

Numerical experiments in the script {\tt csve.m} support this.  The operator $\G$ was discretised on a grid in phase space, with spacing $h=0.3$, bounded by the circle about the origin of radius $b$ for $b=3$, 4 and 5.  Ket space was discretised by taking inner products with the number states $|0〉$ through $|24〉$.  The graph below shows the singular values of the truncated operator $\Gx$.

@@

The contour plot shows how the analytic expression for $σ_n$ depends on $n$ and the truncation radius $b$.  The value has decayed to 0.7 when $n=b²$.  The rule of thumb is that we get one large singular value for every unit circle of phase space, with area $π$.

\topinsert \XeTeXpicfile svcont.pdf width \hsize \endinsert

The following three figures show the right singular vectors, sampled versions of phase space functions.  The brightness of the plots indicates $|v_n(α)|$, the colors indicate phase.  These look very much like truncated number states.  As the grid expands, the pre-images of higher number states fit in it, and more singular values have converged.

\topinsert \hbox{\XeTeXpicfile rsv3.pdf width 0.5\hsize
\XeTeXpicfile rsv4.pdf width 0.5\hsize}
\XeTeXpicfile rsv5.pdf width 0.5\hsize\hfil\break
The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase space is truncated at $|α|=3$, 4 and 5.  These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert
